Prospects of detecting gamma-ray emission from galaxy clusters: cosmic rays and dark matter annihilations

### 2011-06-22 paper.tex

\C{Here, $A_{\rmn{dm}}$ is the particle physics factor and implicitely defined by Equation~\ref{eq:dflux}.}
The different density profiles have some impact on the luminosity from
annihilating DM, although the details of the density profile can be
neglected compared to the dominating boost from substructures
(assuming DM to be cold). In Fig.~\ref{fig:radial_lum} we compare the
radial dependence of the accumulative luminosity from different smooth
cluster density profiles to the boosted luminosity due to substructure
for different mass scales. We recalculate the overdensity, $\rhos$,
for the shallower density profile with $\beta=0.6$ and rescale the
concentration parameter in Eq.~(\ref{eq:cfit}) with $300/160$
\cite{2011ApJ...728L..39N} to account for the more centralized scale
radius in cluster with a shallow inner slope. The emission of this
profile with $\beta=0.6$ is about 30\% larger within $\rvir$ compared
to a cuspy NFW ($\beta=1.0$).  This difference is built up within
$0.1\rvir$ (i.e. close to $r_\s$). Hence the slope of the central part
of a cluster has little influence for the DM luminosity within $\rvir$
as long as the degeneracy in the lensing mass measurements with $r_\s$
(which decreases for decreasing inner slope) has been taken into
account. The emission from an Einasto density profile is about 50\%
larger than the cuspy NFW profile in the periphery of the cluster,
where the difference is mainly built up at a few percent of
$\rvir$. The increase in luminosity due to substructures is neglible
in the center of halos, but integrated out to $\rvir$ the boost
relative to the smooth emission profile amounts to approximately 20
for dwarf galaxies, $200$ for galaxies, and $10^3$ for galaxy
clusters. These boost factors are only realized for direct
annihilation emission (continuum emission or final state radiation) or
IC scattering off homogeneous seed photon fields (CMB). For IC
scattering of SD photons, the overlap of final state leptons and SD
photons is smaller which causes the substructure boost over the smooth
emission to be reduced by roughly two orders of magnitudes. \C{It
  should also be noted that the substructure boosted fluxes from
  clusters are much more extended than for dwarf galaxies, hence more
  difficult to detect with Cherenkov telescopes. In fact, it was shown
  in Sanchez-Conde et al. \cite{2011arXiv1104.3530S} that inside
  $0.1$~deg, where the sensitivity of Cherenkov telescopes is
  maximized, the expected DM flux including substructures from the
  brightest dwarf galaxy is about an order magnitude higher than the
  brightest cluster. However, that work assumed a substructure boost
  of about 50 in the most massive cluster halos which is more than an
  order magnitude smaller than what we use in this work, hence in
  projection we expect this difference to be a factor few smaller.}

@@

The electrons and positrons loose their energy on a timescale that is
shorter than the diffusive timescale in the ICM of galaxy clusters for
cosmic ray electrons which is larger than the Hubble time in our
energy range \cite{1997ApJ...487..529B,2011A&A...527A..99E}. Hence, we
neglect the first term of the r.h.s. in Eq.~(\ref{eq:CRdiffusion}),
and derive an expression for the equilibrium number density:
\begin{eqnarray}
{\frac{\dd n_\e}{\dd \ee}}\left(\ee, r \right) &=&
 \frac{1}{b(\ee, r)} \int_{\ee}^{\mx c^2} \dd \ee' \, 
  q_\e(\ee', r),
\label{eq:nds}\\
b(\ee,r) &=& \tilde{b}
\left[\frac{B^2_{\rm CMB}}{8\pi}+\frac{B^2(r)}{8\pi}+u_\sd(r)\right] \ee^2\,,
\\\tilde{b}&=&\frac{4\sigma_{\rm T}c}{3(m_{\rm e}c^2)^2}\,.
\end{eqnarray}
Here we include the three main radiative loss processes for the cosmic
ray electrons and positrons: (1) IC losses on CMB photons with the
equivalent field strength of the CMB of $B_{\rm
  CMB}=3.24\mu\rmn{G}\,(1+z)^2$, where $z$ is the cosmological
redshift. (2) Synchrotron losses on ambient magnetic fields where we
parameterize the magnetic field in the galaxy cluster by $B(r) =
3\mu\rmn{G} \,[n_{\rm e}(r)/n_{\rm e}(0)]^{\alpha_\B}$. We adopt a
magnetic decline of $\alpha_\B=0.7$ in this work which follows from
flux frozen magnetic fields. (3) IC losses on starlight and dust with
an energy density $u_\sd(r)$ given by Eq.~(\ref{eq:u_SD_r}), where
  we outline the derivation in the following.

@@

\C{We compare the calculated fluxes to gamma-ray upper limits (95\%
  c.l.) set by \Fermi-LAT after 18 months of observations
  \cite{2010ApJ...717L..71A}. In particular, to achieve a more
  reliable comparison we adopt, if nothing else is stated, the maximal
  spatially extended limits since the gamma-ray flux from our
  brightest clusters all have an angular extent $>1^\circ$ on the
  sky when the boost from substructures is included. In fact, published
  \Fermi  upper limits are not derived for the kind of extended
  emission that we find with our treatment of substructures, hence these
  \Fermi limits may not be adequate for some of these clusters. This
  and other recent work highlights that the improved substructure
  models imply that \Fermi limits will have to be re-calculated to
  accommodate for the large source extensions. Note, however, that for
  most clusters the assumption of a point source is well justified.}

\C{The flux upper limits, that we compare to, are a function of spectral index
  $\alpha$. However, in the relevant energy range, $0.1-100$~GeV, the spectral
  index varies within $1.5 < \alpha < 3.0$. This changes the photon flux upper
  limits by $<50\%$, with \Fermi-LAT being more sensitive to a hard
  spectrum \cite{2010ApJ...717L..71A}. For simplicity, we adopt the upper limits
  calculated for a spectral index $\alpha=2$ and estimate the change in the
  differential upper limits for each gamma-ray model and energy range using:
\begin{eqnarray}
\frac{\dd F_\gamma}{\dd \eg} &=& K\,\eg^{-\alpha}\,,\quad F_\gamma = 
K\,\eg^{1-\alpha}\,,\nonumber\\
\bra\eg \,F_\gamma\ket &=& K\,
\frac{\displaystyle\int \dd E\, \eg^{1-\alpha}}{\displaystyle\int \dd \eg}
= \frac{K}{2-\alpha}\,
\frac{E_{\gamma,2}^{2-\alpha}-E_{\gamma,1}^{2-\alpha}}{E_{\gamma,2}-E_{\gamma,1}}\,,\nonumber\\
%\bra\eg \,F_\gamma\ket &=& K\,\frac{\sum \dd \eg^{2-\alpha}}{\sum \dd \eg}
%= K\,\frac{\sum \dd \eg^{2-\alpha}}{E_{\gamma,2}-E_{\gamma,1}}\,,\nonumber\\
\label{eq:spec_ind_UL}
\end{eqnarray}
where $E_{\gamma,1}$ and $E_{\gamma,2}$ are the upper and lower energy
limits, respectively. In the 1-10~GeV energy range we find that the
upper limits for $($LP,BM,CR$)$ change by a factor
$(1.10,\,0.71,\,1.15)$ where we have used the spectral indices
$(2.20,\,1.22,\,2.24)$, respectively. See
Table~\ref{tab:spectral_index} for additional spectral indices.}

@@

\begin{figure*}
\begin{minipage}{2.0\columnwidth}
 \includegraphics[width=0.49\columnwidth]{figures/flux.cluster.Fornax.v13.0.1deg.1.6T.SubMass.SF700.IR2.noMW.woGal.eps}
\includegraphics[width=0.49\columnwidth]{figures/flux.cluster.Coma.v13.0.1deg.1.6T.SubMass.SF700.IR2.noMW.woGal.eps}
\includegraphics[width=0.49\columnwidth]{figures/flux.cluster.Virgo.v13.0.1deg.1.6T.SubMass.SF700.IR2.noMW.woGal.eps}
\includegraphics[width=0.49\columnwidth]{figures/flux.cluster.Perseus.v13.0.1deg.1.6T.SubMass.SF700.IR2.noMW.woGal.eps}
\caption{\it Comparing the gamma-ray emission from different
  clusters. We show the differential flux of clusters \C{in the upper
    panel of each figure}: Fornax (upper left), Coma (upper right),
  Virgo (lower left), and Perseus (lower right). We show CR induced
  emission (red solid), a leptophilic (LP) model that includes both
  final state radiation and IC upscattered CMB, dust and starlight
  (green dash-dotted), and the benchmark $\Kp$ model (BM) that
  includes continuum emission, and IC upscattered CMB, dust and
  starlight (blue dashed). The arrows show the spatially extended
  differential upper limits set by \Fermi-LAT in the energy ranges
  $0.2-1$~GeV, $1-10$~GeV, and $10-100$~GeV from left to right. We
  show the individual components for the BM model and LP model for
  Virgo and Perseus, respectively. \C{In the lower panel of each
    figure we show differential spectral indices, $\alpha_\gamma$,
    where $\dd F_\gamma/\dd E_\gamma \sim E_\gamma^{-\alpha_\gamma}$.}
  The flux from the clusters is integrated out to $\rvir$ using a
  point spread function of $0.1\deg$ and includes the boost from
  substructures. We find that the lower GeV energy regime is most
  constraining due to the peak sensitivity of \Fermi at these energies
  and use this regime to get upper limits on boost factors.}
 \label{fig:clu_comp}
\end{minipage}
\end{figure*}

\C{In the lower panels of Fig.~\ref

For consistency we provide in addition the differential spectral
  indices $\alpha_\gamma=-\frac{\dd \log(\dd F/\dd E)}{\dd \log(E)}$
  for the different emission models in the lower panels of
  Fig.~\ref{fig:clu_comp}. We find very similar spectral indices for
  different clusters when the boost from substructures is substantial,
  while for models without substructures the relative contribution
  from the upscattered dust and starlight breaks the spectral
  universality for the DM models. In the $1~\gev-1~\tev$ energy regime
  the CR spectral index is approximately constant $\sim (2.1-2.3)$,
  while the spectral index for the LP model has a large spread and
  varies between $1.2-4.0$, hence implying that gamma-ray upper limits
  are more sensitive to the specific energy regime for these
  models. Similarly, the index for the BM $\Kp$ model is about 1.0 at
  1~GeV and increases monotonically for higher energies, which also
  suggests that the flux upper limits contrasted to this model are
  sensitive to the specific energy range, and where a small energy
  range would be preferred. In experiments is the spectral index often
  fixed over an energy range, hence we show in
  Tab.~\ref{tab:spectral_index} the indices for our emission models in
  four different energy bands. We find that these spectral indices
  have a much smaller spread, hence reducing the importance of the
  specific energy regimes. For further discussion about the impact of
  different spectral indices on the gamma-ray upper limits see
  Eq.~(\ref{sect:spectral}) and related text.}

\begin{table}
\begin{tabular}{ccccc}
\hline\hline
      Model & $\alpha_{100\,\rmn{MeV}}^{1\,\rmn{GeV}}$  & 
              $\alpha_{1\,\rmn{GeV}}^{10\,\rmn{GeV}}$   & 
              $\alpha_{10\,\rmn{GeV}}^{100\,\rmn{GeV}}$ &
              $\alpha_{100\,\rmn{GeV}}^{1\,\rmn{TeV}}$  \\
\hline
CR-$\pi^0$      & 1.44 & 2.24 & 2.30 & 2.26 \\
DM-LP Sum       & 1.74 & 2.20 & 3.06 & 1.30 \\
DM-BM-$\Kp$ Sum  & 1.28 & 1.22 & 2.23 & -    \\
\hline\hline
\end{tabular}
 \caption{\C{Gamma-ray spectral indices $\alpha_{E_1}^{E_2}$ between
     energies $E_1$ and $E_2$. We show the spectral index for three
     different emission models where the boost from substructures is
     included; the CR induced emission, the leptophilic (LP) model
     that includes both final state radiation and IC upscattered CMB,
     dust and starlight, and the benchmark (BM) $\Kp$ model that
     includes continuum emission, and IC upscattered CMB, dust and
     starlight. The indices are derived for the Fornax cluster,
     although the variance between clusters is very small when the
     enhancement from substructures is included which suppresses the
     relative contribution from the upscattered dust and starlight
     component that is cluster dependent. \label{tab:spectral_index}}}
\end{table}

In this section, we have found that the LP DM models overproduces both
the spatially extended differential and integrated gamma-ray flux
upper limits set by the \Fermi-LAT 18 month data for several
clusters. We can use these upper limits to constrain the boosts due to
SFE and substructures. Furthermore, as more sensitive experiments and
better upper limits emerge, we can start ruling out models that give
rise to boost factors that are manifested in the gamma-ray flux. It is
especially interesting to constrain the SFE, where a boost of the
particle physics cross section of 300 is required to explain the
excess of electrons and positrons observed in the vicinity of Earth by
PAMELA/\Fermi/H.E.S.S. In Fig.~\ref{fig:boost_const} we use the estimated
flux \C{within $\rvir$ of four bright and maximal constraining
  clusters (Fornax, M49, Norma, and Coma, where we have excluded Virgo
  due to its large angular extent) to limit} the SFE as well as the
minimum mass of substructures where
$M_\rmn{lim}\propto\mvir^{0.226}$. We find that M49 and the Fornax
cluster are the most constraining clusters, although the current upper
limits are not good enough to rule out the boost from substructures or
SFE. Assuming that the PAMELA positron excess is due to SFE DM, then
in order not to overproduce the differential upper limits set by
\Fermi-LAT in the energy range\footnote{We average the LP DM flux in
  the energy range $1-10$~GeV before we compare it to the upper
  limits.}  $1-10$~GeV, we can constrain the smallest mass of
substructures to $M_\rmn{lim}\gtrsim 10\,\msun$. Instead assuming the
standard mass of the smallest substructures of $M_\rmn{lim}\approx
10^{-6}\,\msun$, we can constrain the \C{saturated boost from SFE to
  about 14 in M49 and 22 in Fornax, which corresponds to a maximum SFE
  in the MW of about 8 and 13, respectively. Hence, we conclude that
  without substantial improvement in the modelling of these extended
  sources, \Fermi-LAT will not be able to rule out LP models in their
  current form using clusters.} \del{Note that after about six years
  of observations with \Fermi-LAT, we will be able to rule out the LP
  model assuming a substructure boost with a smallest mass of
  $M_\rmn{lim} \approx 10^{-6}\,\msun$.}

@@

\begin{figure}%[t]
 \includegraphics[width=0.99\columnwidth]{figures/SB.resolved.v13.1GeV.SF700.noSuB.vs.SubMass.elmu.eps}
 \caption{\it The impact of substructures on the intrinsic surface
   brightness profile o the Fornax cluster at 1~GeV. The emission
   induced by CRs is denoted by the red solid line, the leptophilic
   model by green lines, and the DM $\Kp$ benchmark model by blue
   lines. The dashed lines show the brightness profiles where the
   boost from substructures is included while the substructures are
   excluded for the dotted lines. The boost from Sommerfeld
   \C{enhancement with and without substructures is 130 and 530
     respectively, where in addition the boost due to substructures is
     about 730.} Notice the nearly flat profiles when substructures
   are included, and the relative large boost at $0.01\rvir$ that is
   due to projection of peripheral substructures onto the cluster
   center.}
 \label{fig:SB_sub}
\end{figure}

@@

\begin{figure*}
\begin{minipage}{2.0\columnwidth}
  \includegraphics[width=0.99\columnwidth]{figures/Fermi.comp.DM.eps}
  \caption{\it \Fermi gamma-ray flux upper limits are contrasted to the
    predicted DM gamma-ray flux. We show the mean differential flux in
    the energy range $E_\gamma=1-10$~GeV for 32 clusters. The
    \C{maximal extended \Fermi-LAT upper limits are shown with black
      arrows, while the point source limits are shown in grey.} The
    predicted fluxes are derived from the dominant inverse
    Compton-scattering of CMB photons in a leptophilic DM model (red),
    and the continuum emission from the DM $\Kp$ benchmark model
    (blue). Assuming a boost factor due to substructures that has a
    mass spectrum extending down to an Earth mass, the expected
    leptophilic fluxes are ruled out by upper limits in 26 of the
    clusters, with the strongest constraints set by Fornax and
    Ophiuchus. We can not yet constrain the benchmark models,
    although improved modelling of extended sources as well as
    stacking of clusters will allow to test these in the upcoming
    years with better data.}
 \label{fig14}
\end{minipage}
\end{figure*}

@@

We identify Fornax to be the best cluster/group candidate for indirect
DM studies as it is expected to emit the brightest annihilation flux
and more importantly, has a comparably low CR induced gamma-ray flux.
\del{This in principle should allow to detect or constrain some BM
  models that have been chosen for the large gamma-ray flux yield with
  three-year \Fermi data (or constrain the substructure boost factor).}
Other nearby bright sources of the extended HIFLUGCS sample of nearby
bright X-ray cluster/groups are M49, Centaurus, and NGC~4636.  The
non-detection of gamma-ray emission by \Fermi in a total of 26
clusters/groups (the mentioned clusters as well as Virgo, Coma, Hydra,
and Norma among others) considerably constrains LP models for DM
annihilation which were introduced to explain the increasing
positron-to-lepton ratio beyond 10 GeV. \C{Assuming that the minimum
  DM substructure mass extends down to an Earth mass, we limit the
  saturated Sommerfeld enhancement factor in M49 and Fornax to about
  14 and 22, respectively. This corresponds to a value in the Milky
  Way (MW) of less than 8 and 13 due to the larger Sommerfeld boost
  factor in smaller mass halos which have a lower velocity dispersion
  that enhances the particle cross section (while assuming
  universality of the DM model).} This would rule out LP models in
  their current form based on the non-observation of gamma-rays in any
  of the fore-mentioned clusters/groups. Alternatively, assuming the
  LP models to be correct, we limit the minimum substructure mass to
  $>10~\msun$ in M49.

>>>>

\C{Here, $A_{\rmn{dm}}$ is the particle physics factor and implicitly defined by Equation~\ref{eq:dflux}.}
The different density profiles have some impact on the luminosity from
annihilating DM, although the details of the density profile can be
neglected compared to the dominating boost from substructures
(assuming DM to be cold). In Fig.~\ref{fig:radial_lum} we compare the
radial dependence of the accumulative luminosity from different smooth
cluster density profiles to the boosted luminosity due to substructure
for different mass scales. We recalculate the overdensity, $\rhos$,
for the shallower density profile with $\beta=0.6$ and rescale the
concentration parameter in Eq.~(\ref{eq:cfit}) with $300/160$
\cite{2011ApJ...728L..39N} to account for the more centralized scale
radius in cluster with a shallow inner slope. The emission of this
profile with $\beta=0.6$ is about 30\% larger within $\rvir$ compared
to a cuspy NFW ($\beta=1.0$).  This difference is built up within
$0.1\rvir$ (i.e. close to $r_\s$). Hence the slope of the central part
of a cluster has little influence for the DM luminosity within $\rvir$
as long as the degeneracy in the lensing mass measurements with $r_\s$
(which decreases for decreasing inner slope) has been taken into
account. The emission from an Einasto density profile is about 50\%
larger than the cuspy NFW profile in the periphery of the cluster,
where the difference is mainly built up at a few percent of
$\rvir$. The increase in luminosity due to substructures is negligible
in the center of halos, but integrated out to $\rvir$ the boost
relative to the smooth emission profile amounts to approximately 20
for dwarf galaxies, $200$ for galaxies, and $10^3$ for galaxy
clusters. These boost factors are only realized for direct
annihilation emission (continuum emission or final state radiation) or
IC scattering off homogeneous seed photon fields (CMB). For IC
scattering of SD photons, the overlap of final state leptons and SD
photons is smaller which causes the substructure boost over the smooth
emission to be reduced by roughly two orders of magnitudes. \C{It
  should also be noted that the substructure boosted fluxes from
  clusters are much more extended than for dwarf galaxies, hence more
  difficult to detect with Cherenkov telescopes. In fact, it was shown
  in Sanchez-Conde et al. \cite{2011arXiv1104.3530S} that inside
  $0.1$~deg, where the sensitivity of Cherenkov telescopes is
  maximized, the expected DM flux including substructures from the
  brightest dwarf galaxy is about an order magnitude higher than the
  brightest cluster. However, that work assumed a substructure boost
  of about 50 in the most massive cluster halos which is more than an
  order magnitude smaller than what we use in this work, hence in
  projection we expect this difference to be a factor few smaller.}

@@

The electrons and positrons loose their energy on a timescale that is
shorter than the diffusive timescale in the ICM of galaxy clusters for
cosmic ray electrons which is larger than the Hubble time in our
energy range \cite{1997ApJ...487..529B,2011A&A...527A..99E}. Hence, we
neglect the first term of the r.h.s. in Eq.~(\ref{eq:CRdiffusion}),
and derive an expression for the equilibrium number density:
\begin{eqnarray}
{\frac{\dd n_\e}{\dd \ee}}\left(\ee, r \right) &=&
 \frac{1}{b(\ee, r)} \int_{\ee}^{\mx c^2} \dd \ee' \, 
  q_\e(\ee', r),
\label{eq:nds}\\
b(\ee,r) &=& \tilde{b}
\left[\frac{B^2_{\rm CMB}}{8\pi}+\frac{B^2(r)}{8\pi}+u_\sd(r)\right] \ee^2\,,
\\\tilde{b}&=&\frac{4\sigma_{\rm T}c}{3(m_{\rm e}c^2)^2}\,.
\end{eqnarray}
Here we include the three main radiative loss processes for the cosmic
ray electrons and positrons: (1) IC losses on CMB photons with the
equivalent field strength of the CMB of $B_{\rm
  CMB}=3.24\mu\rmn{G}\,(1+z)^2$, where $z$ is the cosmological
redshift. (2) Synchrotron losses on ambient magnetic fields where we
parametrize the magnetic field in the galaxy cluster by $B(r) =
3\mu\rmn{G} \,[n_{\rm e}(r)/n_{\rm e}(0)]^{\alpha_\B}$. We adopt a
magnetic decline of $\alpha_\B=0.7$ in this work which follows from
flux frozen magnetic fields. (3) IC losses on starlight and dust with
an energy density $u_\sd(r)$ given by Eq.~(\ref{eq:u_SD_r}), where
  we outline the derivation in the following.

@@

\C{We compare the calculated fluxes to gamma-ray upper limits (95\%
  c.l.) set by \Fermi-LAT after 18 months of observations
  \cite{2010ApJ...717L..71A}. In particular, to achieve a more
  reliable comparison we adopt, if nothing else is stated, the maximal
  spatially extended limits since the gamma-ray flux from our
  brightest clusters all have an angular extent $>1^\circ$ on the sky
  when the boost from substructures is included. In fact, published
  \Fermi upper limits are not derived for the kind of extended
  emission that we find with our treatment of substructures, hence
  these \Fermi limits may not be adequate for some of these
  clusters. This and other recent work highlights that the improved
  substructure models imply that \Fermi limits will have to be
  re-calculated to accommodate for the large source extensions. Note,
  however, that for most clusters the assumption of a point source is
  well justified.  The flux upper limits, that we compare to, are a
  function of spectral index $\alpha$. However, in the relevant energy
  range, $0.1-100$~GeV, the spectral index varies within $1.5 < \alpha
  < 3.0$. This changes the photon flux upper limits by $<50\%$, with
  \Fermi-LAT being more sensitive to a hard spectrum
  \cite{2010ApJ...717L..71A}.}

\del{For simplicity, we adopt the upper limits
  calculated for a spectral index $\alpha=2$ and estimate the change in the
  differential upper limits for each gamma-ray model and energy range using:
\begin{eqnarray}
\frac{\dd F_\gamma}{\dd \eg} &=& K\,\eg^{-\alpha}\,,\quad F_\gamma = 
K\,\eg^{1-\alpha}\,,\nonumber\\
\bra\eg \,F_\gamma\ket &=& K\,
\frac{\displaystyle\int \dd E\, \eg^{1-\alpha}}{\displaystyle\int \dd \eg}
= \frac{K}{2-\alpha}\,
\frac{E_{\gamma,2}^{2-\alpha}-E_{\gamma,1}^{2-\alpha}}{E_{\gamma,2}-E_{\gamma,1}}\,,\nonumber\\
%\bra\eg \,F_\gamma\ket &=& K\,\frac{\sum \dd \eg^{2-\alpha}}{\sum \dd \eg}
%= K\,\frac{\sum \dd \eg^{2-\alpha}}{E_{\gamma,2}-E_{\gamma,1}}\,,\nonumber\\
\label{eq:spec_ind_UL}
\end{eqnarray}
where $E_{\gamma,1}$ and $E_{\gamma,2}$ are the upper and lower energy
limits, respectively. In the 1-10~GeV energy range we find that the
upper limits for $($LP,BM,CR$)$ change by a factor
$(1.10,\,0.71,\,1.15)$ where we have used the spectral indices
$(2.20,\,1.22,\,2.24)$, respectively. See
Table~\ref{tab:spectral_index} for additional spectral indices.}

@@

\begin{figure*}
\begin{minipage}{2.0\columnwidth}
 \includegraphics[width=0.49\columnwidth]{figures/flux.cluster.Fornax.v13.0.1deg.1.6T.SubMass.SF700.IR2.noMW.woGal.eps}
\includegraphics[width=0.49\columnwidth]{figures/flux.cluster.Coma.v13.0.1deg.1.6T.SubMass.SF700.IR2.noMW.woGal.eps}
\includegraphics[width=0.49\columnwidth]{figures/flux.cluster.Virgo.v13.0.1deg.1.6T.SubMass.SF700.IR2.noMW.woGal.eps}
\includegraphics[width=0.49\columnwidth]{figures/flux.cluster.Perseus.v13.0.1deg.1.6T.SubMass.SF700.IR2.noMW.woGal.eps}
\caption{\it Comparing the gamma-ray emission from different
  clusters. We show the differential flux of clusters \C{in the upper
    panel of each figure}: Fornax (upper left), Coma (upper right),
  Virgo (lower left), and Perseus (lower right). We show CR induced
  emission (red solid), a leptophilic (LP) model that includes both
  final state radiation and IC upscattered CMB, dust and starlight
  (green dash-dotted), and the benchmark $\Kp$ model (BM) that
  includes continuum emission, and IC upscattered CMB, dust and
  starlight (blue dashed). The arrows show the spatially extended
  differential upper limits set by \Fermi-LAT in the energy ranges
  $0.2-1$~GeV, $1-10$~GeV, and $10-100$~GeV from left to right
  \C{(using Gau{\ss} profiles for the source extension with
    $\sigma=\{1.2,1.0,0.8\}$ for \{Virgo, Coma, Fornax\} and an
    X-ray-inferred King profile for Perseus)
    \cite{2010ApJ...717L..71A}}. We show the individual components for
  the BM model and LP model for Virgo and Perseus, respectively. \C{In
    the lower panel of each figure we show differential spectral
    indices, $\alpha_\gamma$, where $\dd F_\gamma/\dd E_\gamma \sim
    E_\gamma^{-\alpha_\gamma}$.}  The flux from the clusters is
  integrated out to $\rvir$ using a point spread function of $0.1\deg$
  and includes the boost from substructures. We find that the lower
  GeV energy regime is most constraining due to the peak sensitivity
  of \Fermi at these energies and use this regime to get upper limits
  on boost factors.}
 \label{fig:clu_comp}
\end{minipage}
\end{figure*}

\C{In the lower panels of Fig.~\ref{fig:clu_comp}, we also provide the
  differential spectral index, $\alpha_\gamma=-\dd \log(\dd F/\dd
  E)/\dd \log(E)$, for the different emission models. We find very
  similar spectral indices for different clusters assuming a
  substantial boost from substructures, while for models without
  substructures the relative contribution from the upscattered dust
  and starlight breaks the spectral universality for the DM models. In
  the $1~\gev-1~\tev$ energy regime the CR spectral index is
  approximately constant $\sim (2.1-2.3)$, while the spectral index
  for the LP model varies substantially between $1.2-4.0$, hence
  implying that gamma-ray upper limits are more sensitive to the
  specific energy regime for these models. Similarly, the index for
  the BM $\Kp$ model is about 1.0 at 1~GeV and increases monotonically
  toward higher energies. To maximize photon count statistics,
  experiments calculate band-integrated fluxes which implies a fixed
  spectral index over that energy range. Hence, we show $\alpha_\gamma$ in
  Table~\ref{tab:spectral_index} for our emission models in four
  different energy bands. We find smaller variations for the banded
  $\alpha_\gamma$-values, hence reducing the importance of the specific
  energy regimes.}

\del{ For further discussion about the impact of different
  spectral indices on the gamma-ray upper limits see
  Eq.~(\ref{sect:spectral}) and related text.}

\begin{table}
\begin{tabular}{ccccc}
\hline\hline
      Model & $\alpha_{100\,\rmn{MeV}}^{1\,\rmn{GeV}}$  & 
              $\alpha_{1\,\rmn{GeV}}^{10\,\rmn{GeV}}$   & 
              $\alpha_{10\,\rmn{GeV}}^{100\,\rmn{GeV}}$ &
              $\alpha_{100\,\rmn{GeV}}^{1\,\rmn{TeV}}$  \\
\hline
CR-$\pi^0$      & 1.44 & 2.24 & 2.30 & 2.26 \\
DM-LP Sum       & 1.74 & 2.20 & 3.06 & 1.30 \\
DM-BM-$\Kp$ Sum  & 1.28 & 1.22 & 2.23 & -    \\
\hline\hline
\end{tabular}
\caption{\C{Banded gamma-ray spectral index $\alpha_{E_1}^{E_2}$ between
    energies $E_1$ and $E_2$. We show the spectral index for three
    different emission models where the boost from substructures is
    included; the CR induced emission, the leptophilic (LP) model
    that includes both final state radiation and IC upscattered CMB,
    dust and starlight, and the benchmark (BM) $\Kp$ model that
    includes continuum emission, and IC upscattered CMB, dust and
    starlight. $\alpha_\gamma$-values are derived for the Fornax cluster,
    although the variance between clusters is very small when we account for the
    enhancement from substructures. This  suppresses the
    relative contribution from the upscattered dust and starlight
    component that depends on  cluster mass. \label{tab:spectral_index}}}
\end{table}

In this section, we have found that the LP DM models overproduces both
the spatially extended differential and integrated gamma-ray flux
upper limits set by the \Fermi-LAT 18 month data for several
clusters. We can use these upper limits to constrain the boosts due to
SFE and substructures. Furthermore, as more sensitive experiments and
better upper limits emerge, we can start ruling out models that give
rise to boost factors that are manifested in the gamma-ray flux. It is
especially interesting to constrain the SFE, where a boost of the
particle physics cross section of $\sim$300 is required to explain the
excess of electrons and positrons observed in the vicinity of Earth by
PAMELA/\Fermi/H.E.S.S. In Fig.~\ref{fig:boost_const} we use the estimated
flux \C{within $\rvir$ of four bright and maximal constraining
  clusters (Fornax, M49, Norma, and Coma, where we have excluded Virgo
  due to its large angular extent) to limit} the SFE as well as the
minimum mass of substructures where
$M_\rmn{lim}\propto\mvir^{0.226}$. We find that M49 and the Fornax
cluster are the most constraining clusters, although the current upper
limits are not good enough to rule out the boost from substructures or
SFE. Assuming that the PAMELA positron excess is due to SFE DM, then
in order not to overproduce the differential upper limits set by
\Fermi-LAT in the energy range\footnote{We average the LP DM flux in
  the energy range $1-10$~GeV before we compare it to the upper
  limits.}  $1-10$~GeV, we can constrain the smallest mass of
substructures to $M_\rmn{lim}\gtrsim 10\,\msun$. Instead, assuming the
standard mass of the smallest substructures of $M_\rmn{lim}\approx
10^{-6}\,\msun$, we can constrain the \C{saturated boost from SFE to
  about 14 in M49 and 22 in Fornax, which corresponds to a maximum SFE
  in the MW of about 8 and 13, respectively. Hence, we conclude that
  without substantial improvement in the modelling of these extended
  sources, \Fermi-LAT will not be able to rule out LP models in their
  current form using clusters.} \del{Note that after about six years
  of observations with \Fermi-LAT, we will be able to rule out the LP
  model assuming a substructure boost with a smallest mass of
  $M_\rmn{lim} \approx 10^{-6}\,\msun$.}

@@

\begin{figure}%[t]
 \includegraphics[width=0.99\columnwidth]{figures/SB.resolved.v13.1GeV.SF700.noSuB.vs.SubMass.elmu.eps}
 \caption{\it The impact of substructures on the intrinsic surface
   brightness profile o the Fornax cluster at 1~GeV. The emission
   induced by CRs is denoted by the red solid line, the leptophilic
   model by green lines, and the DM $\Kp$ benchmark model by blue
   lines. The dashed lines show the brightness profiles where the
   boost from substructures is included while the substructures are
   excluded for the dotted lines. The boost from Sommerfeld
   \C{enhancement with and without substructures is 130 and 530
     respectively. The boost due to substructures is
     about 730.} Notice the nearly flat profiles when substructures
   are included, and the relative large boost at $0.01\rvir$ that is
   due to projection of peripheral substructures onto the cluster
   center.}
 \label{fig:SB_sub}
\end{figure}

@@

\begin{figure*}
\begin{minipage}{2.0\columnwidth}
  \includegraphics[width=0.99\columnwidth]{figures/Fermi.comp.DM.eps}
  \caption{\it \Fermi gamma-ray flux upper limits are contrasted to 
    predicted DM gamma-ray fluxes. We show the mean differential flux in
    the energy range $E_\gamma=1-10$~GeV for 32 clusters. The
    \C{maximally extended \Fermi-LAT upper limits are shown with black
      arrows, while the point source limits are shown in grey.} The
    predicted fluxes are derived from the dominant inverse
    Compton-scattering of CMB photons in a leptophilic DM model (red),
    and the continuum emission from the DM $\Kp$ benchmark model
    (blue). Assuming a boost factor due to substructures that has a
    mass spectrum extending down to an Earth mass, the expected
    leptophilic fluxes are ruled out by upper limits in 26 of the
    clusters, with the strongest constraints set by Fornax and
    Ophiuchus. We can not yet constrain the benchmark models,
    although improved modelling of extended sources as well as
    stacking of clusters will allow to test these in the upcoming
    years with better data.}
 \label{fig14}
\end{minipage}
\end{figure*}

@@

We identify Fornax to be the best cluster/group candidate for indirect
DM studies as it is expected to emit the brightest annihilation flux
and more importantly, has a comparably low CR induced gamma-ray flux.
\del{This in principle should allow to detect or constrain some BM
  models that have been chosen for the large gamma-ray flux yield with
  three-year \Fermi data (or constrain the substructure boost
  factor).}  Other nearby bright sources of the extended HIFLUGCS
sample of nearby bright X-ray cluster/groups are M49, Centaurus, and
NGC~4636.  The non-detection of gamma-ray emission by \Fermi in a
total of 26 clusters/groups (the mentioned clusters as well as Virgo,
Coma, Hydra, and Norma among others) considerably constrains LP models
for DM annihilation which were introduced to explain the increasing
positron-to-lepton ratio beyond 10 GeV. \C{Assuming that the minimum
  DM substructure mass extends down to an Earth mass and that the DM
  annihilation flux is dominated by small scale substructure, we limit
  the saturated Sommerfeld enhancement factor in M49 and Fornax to
  about 14 and 22, respectively. This corresponds to a value in the
  Milky Way (MW) of less than 8 and 13, respectively, due to the larger Sommerfeld
  boost factor in smaller mass halos which have a lower velocity
  dispersion that enhances the particle cross section (while assuming
  universality of the DM model).} This would rule out LP models in
their current form based on the non-observation of gamma-rays in any
of the fore-mentioned clusters/groups. Alternatively, assuming the LP
models to be correct, we limit the minimum substructure mass to
$>10~\msun$ in M49.
